---
title: "Reproducible Research: Peer Assessment 1"
author: "Pierre_Lecointre"
date: "Friday, November 14, 2014"
output: html_document
keep_md: true
---
## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit][1], [Nike Fuelband][2], or [Jawbone Up][3].  
These type of devices are part of the "quantified self" movement -- a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day.  
The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

Let's prepare our environment :
```{r setoptions}
# to be able to read all code chunks, and place plots in the expected folder
library(knitr)
opts_chunk$set(echo=TRUE, fig.path="figures/")
# libraries used
library(ggplot2)
```

## Loading and preprocessing the data
Data are stored in an archive file (activity.zip), in a comma separated format (csv).

### 1. Load the data
```{r loaddata}
unzip("activity.zip")
unlink("activity.zip")
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
```
### 2. Process / transform the data
```{r preprocess}
# Change class for the date variable
activity$date <- as.Date(activity$date)
```
Each observation contains 3 variables :

* the number of steps (`steps`); missing values coded as `NA`
* the date (`date`) in YYYY_MM-DD format
* a label for the 5 minute interval (`interval`) in an hhmm format (hh : hour, 0 to 23; mm : minutes, 0 to 55 in a 5 min steps).

Let's have a look to a sample of the dataset : 
```{r}
head(activity)
```

## What is mean total number of steps taken per day?

For this part, we ignore the mission values in the dataset.

### 1.make a histogram of the total number of steps taken each day.
```{r}
# Aggregate data by day
activity_day <- aggregate(steps~date, data = activity, FUN = sum, na.rm=TRUE)
# Plot histogram with ggplot2
histogram <- ggplot (activity_day, aes(x=date, y=steps)) +
        geom_bar() +
        labs(title='Figure 1: Number of steps taken daily\n',y='Total steps per day', x='Date')
plot(histogram)

```



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?


[1]:http://www.fitbit.com/ "Fitbit"
[2]:http://www.nike.com/us/en_us/c/nikeplus-fuelband/ "Nike Fuelband"
[3]:https://jawbone.com/up/ "Jawbone Up"
